\pagebreak
{\Huge \textbf{Acknowledgements}}
\vspace{6mm}

After three years of work spent in the development of this Ph.D. thesis,
many steps toward the integration between the localization technique and the
n-electron valence state perturbation theory have been made, paving the way
for a powerful tool in applied quantum chemistry. Many other steps lie
ahead, much more integration and coordination is needed, but in the long
term every bit of knowledge will be integrated in a common, reliable and
powerful meta-code, an important platform for theoretical and experimental
investigations.

Chemistry is everywhere. New drugs and insights in biological processes, new
materials for improving storage, computation, communication in a fast
changing world. In developing new theories and computational approaches, we
must keep in mind that the very high potential of these new tools still has
to be completely unleashed. Once this happens, we will be able to perform a
new kind of research.

I wish to thank my supervisors Prof. Renzo Cimiraglia, Prof. Stefano
Evangelisti and also Dr. Daniel Maynau and Dr. Celestino Angeli for
providing the passionate environment needed to develop a satisfying
research.

I wish to thank Dr. Elda Rossi, Dr. Antonio Monari and Prof. Gianluigi
Bendazzoli for the useful discussions to produce Q5Cost and F90xml
libraries.

I wish to thank Dr. Mariachiara Pastore for the strong moral support and
help during the Ph.D.

I wish to thank Prof. Maurizio Dal Colle for the help and hospitality.

I wish to thank Dr. Alex Cavallini, Dr. Manuel Sparta, Dr. Mirko Cestari,
Dr. Lara Ferrighi for the fantastic research experience.

I wish to thank Prof. Gilli, for allowing me to discover new insights in my
career.

Finally, I wish to thank my Family, Paola Marani and Walter Padovani, for
the help in achieving all the professional steps needed to obtain this Ph.D.
