\pagestyle{fancy}
\chapter{Grid integration}
\label{chp:grid}
%\pagenumbering{arabic}

Computational power is cheap. Storage is also cheap and networking as well.
A well known empirical law for the increase of processing power is the Moore's law
\cite{moore-law}, which dictates the doubling of computational speed every
18 months. Similar laws have been envised for networking (doubling every 9
months) and storage capacity (doubling every 12 months). For scientific
needs, computational power and storage often are the limiting factors
for new developments and insights. 

The dramatic increase of networking performances opens new possibilities to
obtain computational power and storage from unused or underutilized
resources.  This deploys the fundation of a new paradigm for distributed
computing: grid metasystems.  Using grids, it is possible to pool
\begin{itemize}
\item \textit{Fabric resources}, like disk space, computational power, RAM storage,
data aggregation and archiviation, instruments
\item \textit{Logical resources}, like experimentation, data analysis, modeling,
simulation
\item \textit{Human resources}, like competence sharing, administrative burden,
communication.
\end{itemize}
The result is a distributed, heterogeneous metasystem, which provides a
virtual structure to the end user.  Interaction with services is
standardized and uniform, regardless of differences like CPU architecture
or geographical location. In this panorama, a grid provides
virtually infinite resources ready to use to produce a better science.

The most famous example of a grid is probably the Web. The well-known WWW is
a very simple storage grid where documents and informations are
geographically distributed. Users can access these informations through a
well-known interface (a web browser) which makes use of a low-level
uniform protocol (HTTP\cite{http-site}) in order to access resources.
The abstraction provided by the Apache webserver\cite{apache-site}
translates the HTTP requests into fabric level access, like reading or
writing a given file, or running a script.

Consortiums like IETF\cite{ietf-site}, W3C\cite{w3c-site} and
OASIS\cite{oasis-site} ratified standards in order to define
guidelines from low-level issues to high-level protocols and interactions:
TCP/IP\cite{tcp-site,ip-site}, HTTP, XML\cite{xml-site},
SOAP\cite{soap-site}, WSDL \cite{wsdl-site}, UDDI \cite{uddi-site}, RDF
\cite{rdf-site}, WSFL \cite{wsfl-site} are some examples of these standards.
Toolkits, frameworks and utilities have been developed, both commercial and
opensource to exploit the grid paradigm: Legion \cite{legion-site}, Globus
\cite{globus-site,ogsa-site}, and Unicore \cite{unicore-site} are some examples.

At the moment, a restricted set of grid applications can be devised:
\begin{itemize}
\item \textit{Low communication applications}: also known as ``embarassingly
parallel'' applications. For these applications, the problem is decomposed in
small almost independent pieces (usually called \textit{Working Units}),
and the computation is carried over each working unit. This category is
well represented by Seti@Home\cite{seti-site} or weather analysis
\item \textit{Staged applications}: the procedure is decomposed in
serialized steps, each one performed on a given site on the grid. This
approach is useful to manage human factors, like competence sharing and
administrative burden, license problems, use of private code, security
reasons
\item \textit{Resource access}: a resource from a service provider is made
accessible by request. Classic examples could be databases, instruments,
storage farms.
\end{itemize}

For quantum chemical processes some experience has been made, mainly with
parallelization of various algorithms (using PVM \cite{pvm-site}, MPI \cite{mpi-site} or OpenMP
\cite{openmp-site}). However, quantum chemistry evaluations are mainly
serialized in nature: for \textit{ab initio} calculations, a step involving
the evaluation of integrals is followed by other treatments, which in turn
provide information for others and so on.  This calls for a ``Staged
application'' architecture. This architecture is also enforced by
administrative issues, competence sharing and license problems, and finally
a current intrinsic complexity in the creation of multiplatform codes.

\input{04_grid/text_metachem_grid.tex}
\input{04_grid/text_q5cost_introduction.tex}
\input{04_grid/text_q5cost_implementation.tex}
\input{04_grid/text_f77xml_introduction.tex}
\input{04_grid/text_f77xml_implementation.tex}

