\pagestyle{fancy}
\chapter{Localization}
\label{chp:localization}
%\pagenumbering{arabic}
\begin{center}
\textit{This chapter, in particular for the computational evaluation,
is based on the article at Ref. \citen{jcc-26-1042-2005}}
\end{center}
{\ }\\
\vspace{-1mm}

The need to evaluate the properties of large molecular systems using a high-level
\textit{ab initio} approach is an interesting and challenging task for the
quantum chemist. Many useful results have been achieved in this direction,
thanks to refinements of theoretical approaches, together with computational
resources increasingly more efficient and cheap.  However, the problem of
the growth of the computational effort as a function of the system size
still represents, in most cases, a serious bottleneck.  This can lead to a
huge computational cost even for relatively small systems.  

Many techniques have been implemented to deal with large systems, like, for
instance, metallic complexes with organic ligands, or biologically
interesting molecules.  For example, in the evaluation of potential energy
surfaces of the ground state a Molecular Mechanics approach makes feasible
calculations on large molecular systems. The price to pay, in this case, is
to consider the system as purely classical.

When a quantum mechanical evaluation is needed (description of excited states,
bond breaking, etc.), the mixing of molecular mechanics with
quantum mechanics is a possible strategy, like Quantum Mechanics/Molecular
Mechanics QM/MM \cite{jcc-11-700-1990} and the ``Our own N-layered Integrated
molecular Orbital and molecular Mechanics'' ONIOM method
\cite{jpc-100-19357-1996}. For the calculation of excited states,
Time Dependent Density Functional Theory (TDDFT) methods are also
available, which behave well in some cases but with substantial errors in
others, as pointed out for example in Ref. \citen{jacs-126-4007-2004}. 

Performing purely \textit{ab initio} evaluations is also feasible. A
possible approach is represented by the elimination of large molecular
groups, replacing them by computationally cheaper mimicking entities, using
Effective Group Potential (EGP) pseudo-potentials
\cite{jpca-105-198-2001,jpca-105-206-2001}, for example, or simply by
hydrogen atoms. Other available methods, like the Symmetry Adapted
Cluster-Configuration Interaction
(SAC-CI)\cite{jcp-68-2053-1978,cpl-59-362-1978,cpl-67-329-1979,cpl-67-334-1979}
have been applied successfully to large biological systems
\cite{cpl-250-437-1996}.

Another possible strategy in the treatment of large molecular systems is
given by Localized Molecular Orbitals (LMO). The use of localized orbitals in
quantum chemistry has a long history \cite{chavet-ladiqc}.  Only in recent
years, however, Localized Molecular Orbitals have become popular as a key
step toward Linear Scaling \textit{ab initio} approaches, where the
computational cost of the evaluation grows linear with respect to the
molecular system size. Using a localized philosophy, a reduction of the
high computational cost imposed by delocalized methods can be obtained.

\input{02_localization/01_delocalization_and_localization}
\input{02_localization/02_localization_technique}
\input{02_localization/03_freeze_and_cut_theory}
\input{02_localization/04_freeze_and_cut_implementation}
\pagebreak
\section{Evaluations}
\input{02_localization/05_example1_ammoniotridecenoate}
\input{02_localization/06_example2_tridequenal}
\input{02_localization/07_example3_acetone_water}
\input{02_localization/08_conclusions}
